---
title: "Expression of carboxylesterase and paraoxonase in the placenta and their association with chlorpyrifos exposure during pregnancy"
author: "Piuque M. Rodriguez, Berta Vera, Carolina Burgos, Gustavo Gimenez, Karina 
Miglioranza, Cristina L. Ramirez, Andrea Lavalle , Paola M. Ondarza, Natalia L.
Guiñazú"
format: 
  html:
    toc: true
    toc-depth: 3
editor: visual
---

# Prepare the data

# Load essential libraries for data manipulation


Loading basic libraries to manipulate data.


```{r}
library(tidyverse) #tidyverse environment
```




# Import data from .ods file


```{r}
datos_clor <- readODS::read_ods("Tabla_05_04_25.ods")
```



# Display basic summary statistics


```{r}
summary(datos_clor)
```


Look some NA values.



An descripción of columns
```{r}
glimpse(datos_clor)
```


First past to factor all character columns



```{r}
datos_clor <-  datos_clor |>   mutate(
    Niv_educ = factor(Niv_educ,
                      levels = 1:4,
                      labels = c("No education/Primary incomplete",
                                 "Primary complete",
                                 "Secondary complete",
                                 "Greater")),
    Sexo_neo = factor(Sexo_neo,
                      levels = c(0, 1),
                      labels = c("Femenino", "Masculino")),
    Fumadora_pas = factor(Fumadora_pas,
                      levels = c(0, 1),
                      labels = c("NO", "SI")),
     Fumiga = factor(Fumiga,
                      levels = c(0, 1),
                      labels = c("NO", "SI")),
    Pozo = factor(Pozo,
                      levels = c(0, 1),
                      labels = c("NO", "SI")),
    Chlorpyrifos_detection = case_when(
      Chlorpyrifos_detection == ">LOQ" ~ 1L,
      Chlorpyrifos_detection == "<LOD" ~ 0L,
      TRUE ~ NA_integer_
    ),
    Paridad = as.integer(Paridad),
    IMC = factor(IMC,
                      levels = 0:3,
                      labels = c("normal",
                                 "underweight",
                                 "overweight",
                                 "obesity")),
    Grupo = fct_relevel(Grupo, "Urbano", "Rural")
  )

datos_clor

```



 Count NA values for each column in long format

```{r}
datos_clor  |> 
  summarise(across(everything(), ~sum(is.na(.)))) |> 
  pivot_longer(cols = everything(), names_to = "columna", values_to = "n_NA") |> 
  mutate(tipo = sapply(datos_clor[, columna], class)) %>%
  relocate(tipo, .after = columna) %>%
  arrange(desc(n_NA))

```


Compute means in some variables.


```{r}
datos_clor  |> group_by(Grupo) |> summarise(Medias = mean( Ind_Pond, na.rm = TRUE))
```


# Compute Pearson correlation matrix with complete cases only



```{r}
matr_corr <- datos_clor |> select(where(is.numeric)) |> cor(use = "complete.obs") |> round(digits = 2)
```



# Compute Spearman correlation matrix with complete cases only


```{r}
datos_clor |> select(where(is.numeric)) |> cor(use = "complete.obs", method = "spearman") |> round(digits = 2)
```



# Visualize correlation matrix with corrplot


```{r, fig.width=8.5, fig.height=8.5}
library(corrplot)


# Create the correlogram 
corrplot(matr_corr, type = "upper",  
         method = "square",  
         addCoef.col = "black",  
         tl.col = "black", 
         tl.srt = 45,
         tl.cex = 0.8,      # Tamaño de letra
         number.cex = 0.7)  # Ta)
```



Plot pairwise relationships with GGally::ggpairs



```{r warning=FALSE, fig.align='center', fig.width=7, fig.height=6}
library(GGally)
datos_clor  |>   ggpairs(    # Data frame
        columns = c(14:20,27)) # Columns
```







```{r warning=FALSE, fig.align='center', fig.width=7, fig.height=6}
library(GGally)
datos_clor  |>   ggpairs(    # Data frame
        columns = c(14:20,27),
        aes(color = Grupo,  # Color by group (cat. variable)
            alpha = 0.5),
        upper = list(continuous = wrap("cor", size = 2.5))
        )
         # Columns
```



# Fit linear model to CPF


```{r}
modelo_lm <- lm(CPF ~ Edad +Pozo+IMC+Fumiga+Niv_educ+Paridad+Fumadora_pas+
                Edad_Gest+Sexo_neo+CES+CES_4_MUBA+CES2+PON+PON_DHC+
                  PON2+Peso_Neo+Peso_Placenta+Talla+Per_Cef+Ind_Plac+Ind_Pond     , 
                data = datos_clor)
summary(modelo_lm)
```



```{r}
car::Anova(modelo_lm, type="III")
```


No found any effect since F test was not significatn.


# Fit logistic regression model to binary variable Chlorpyrifos_detection


```{r}
modelo_glm <- glm(Chlorpyrifos_detection ~ Edad +Pozo+IMC+Fumiga+Niv_educ+Paridad+Fumadora_pas+
                Edad_Gest+Sexo_neo+CES+CES_4_MUBA+CES2+PON+PON_DHC+
                  PON2+Peso_Neo+Peso_Placenta+Talla+Per_Cef+Ind_Plac+Ind_Pond + Grupo    , 
                data = datos_clor, family = binomial(link = "logit"))
summary(modelo_glm)
```


Clean the NA's values. Remove rows with na values.Perform stepwise selection using BIC


```{r}
datos_clor_limpio <- datos_clor %>% drop_na()
modelo_glm_limpio <- glm(Chlorpyrifos_detection ~ Edad +Pozo+IMC+Fumiga+Niv_educ+Paridad+Fumadora_pas+
                Edad_Gest+Sexo_neo+CES+CES_4_MUBA+CES2+PON+PON_DHC+
                  PON2+Peso_Neo+Peso_Placenta+Talla+Per_Cef+Ind_Plac+Ind_Pond + Grupo    , 
                data = datos_clor_limpio, family = binomial(link = "logit"))
modelo_seleccion <- step(modelo_glm_limpio, direction = "both", k = log(63))
summary(modelo_seleccion)

```


Only two variables was significant Ind_Pond and Group. Test posible interaction or aditive effects.


```{r}

modelo_final1 <- glm(formula = Chlorpyrifos_detection ~ Ind_Pond * Grupo, 
                    family = binomial(link = "logit"),     data = datos_clor_limpio)
modelo_final2 <- glm(formula = Chlorpyrifos_detection ~ Ind_Pond + Grupo, 
                    family = binomial(link = "logit"),    data = datos_clor_limpio)
modelo_final3 <- glm(formula = Chlorpyrifos_detection ~  Grupo, 
                    family = binomial(link = "logit"),     data = datos_clor_limpio)
anova(modelo_final1, modelo_final2, modelo_final3, test = "LRT")

```


There was aditive efect only.Then check overdispersion in the model.


```{r}
summary(modelo_final2)
performance::check_overdispersion(modelo_final2)
```


Extract coefficients. In this case logODDs.


```{r}
summary(modelo_final2)$coefficients
```



Transform to Odds Ratios by means log transformation.



```{r}
exp(coefficients(modelo_final2))[2:3]
```

Compute Pseudo R2.

```{r}
DescTools::PseudoR2( modelo_final2, which = "all")
```


Predict probabilities to mean ind_pon and Rural Group.


```{r}
newdata <- data.frame(Ind_Pond = 2.96, Grupo = "Rural")
predict(modelo_final2, newdata = newdata, type = "response")

```


The probability to detect cp is too high in rural group.


```{r}
newdata <- data.frame(Ind_Pond = 2.9, Grupo = "Urbano")
predict(modelo_final2, newdata = newdata, type = "response")

```


We can understand the final model as an ANCOVA. Try to plot them.
```{r}
# Crear un rango de valores para Ind_Pond
x_vals <- seq(min(datos_clor_limpio$Ind_Pond), 
              max(datos_clor_limpio$Ind_Pond), 
              length.out = 100)

# Crear data.frame con combinaciones para Rural y Urbano
nuevos_datos <- data.frame(
  Ind_Pond = rep(x_vals, 2),
  Grupo = rep(c("Rural", "Urbano"), each = length(x_vals))
)



# Predict probabilities and plot logistic model



nuevos_datos$prob <- predict(modelo_final2, newdata = nuevos_datos, type = "response")

# Graficar
library(ggplot2)
ggplot(nuevos_datos, aes(x = Ind_Pond, y = prob, color = Grupo)) +
  geom_line(size = 1.2) +
  labs(
    title = "Probabilidad de detección de Chlorpyrifos",
    x = "Ind_Pond",
    y = "Probabilidad de detección",
    color = "Grupo"
  ) +
  theme_minimal()
```



```{r}
ggplot(nuevos_datos, aes(x = Ind_Pond, y = prob, color = Grupo)) +
  geom_line(size = 1.2) +
  geom_jitter(data = datos_clor_limpio, 
              aes(x = Ind_Pond, y = Chlorpyrifos_detection, color = Grupo),
              height = 0.05, alpha = 0.5) +
  labs(
    title = "Modelo Logístico: Detección de Chlorpyrifos",
    x = "Ind_Pond",
    y = "Probabilidad de detección",
    color = "Grupo"
  ) +
  theme_minimal()

```



# Further reduced models with AIC



```{r}
modelo_glm_limpio <- glm(Chlorpyrifos_detection ~ Edad +Pozo+IMC+Fumiga+Niv_educ+Paridad+Fumadora_pas+
                Edad_Gest+Sexo_neo+CES+CES_4_MUBA+CES2+PON+PON_DHC+
                  PON2+Peso_Neo+Peso_Placenta+Talla+Per_Cef+Ind_Plac+Ind_Pond + Grupo    , 
                data = datos_clor_limpio, family = binomial(link = "logit"))
modelo_seleccionAIC <- step(modelo_glm_limpio, direction = "both")
summary(modelo_seleccionAIC)

```







```{r}
glm_AIC_reducido <- glm(formula = Chlorpyrifos_detection ~     Edad_Gest + CES2 + PON2 + Talla + Ind_Pond + Grupo, family = binomial(link = "logit"), 
    data = datos_clor_limpio)
summary(glm_AIC_reducido)
```

```{r}
glm_AIC_reducido2 <- glm(formula = Chlorpyrifos_detection ~     Edad_Gest + CES2 + Talla + Ind_Pond + Grupo, family = binomial(link = "logit"), 
    data = datos_clor_limpio)
summary(glm_AIC_reducido2)
```



## Log-transformed CPF model



```{r}
modelo_lm2 <- lm(log(CPF) ~ Edad +Pozo+IMC+Fumiga+Niv_educ+Paridad+Fumadora_pas+
                Edad_Gest+Sexo_neo+CES+CES_4_MUBA+CES2+PON+PON_DHC+
                  PON2+Peso_Neo+Peso_Placenta+Talla+Per_Cef+Ind_Plac+Ind_Pond     , 
                data = datos_clor)
summary(modelo_lm2)
```


```{r}
car::Anova(modelo_lm2, type="III")
```


No effect was found
